---
title: "Post Transplant Data Clean"
output:
    html_notebook:
      toc: yes
      toc_depth: 2
      toc_float: yes

---
 
  PROJECT:    	Accuracy of of LAS
  
  PROGRAMMER: 	Nicole Dussault, Will Parker
  
  PROGNAME:   	Post_Transplant_Data_Clean
  
  CREATED:    	7/5/2019
  
  LAST UPDATED: "`r format(Sys.time(), '%d %B, %Y')`"
  
  OBJECTIVE: 	  Create Post-transplant dataset for LAS accuracy analysis   
  
  NOTES:        Updated for Feb 2015 - Feb 2019
  


```{r Load_libraries, include = FALSE}

rm(list=ls())
library(tidyverse)
library(haven)
library(lubridate)
library(survival)
library(ggfortify)

```


```{r Convenience_functions, include = FALSE}

#Convenience functions

numextract <- function(string){
  as.numeric(str_extract(string, "\\-*\\d+\\.*\\d*"))
}

comma <- function(x) format(x, digits = 3, big.mark = ",")

comma_0 <- function(x) format(x, digits = 0, big.mark = ",")


comma_1 <- function(x) format(x, digits = 1, big.mark = ",")

comma_2 <- function(x) format(x, digits = 2, big.mark = ",")

comma_p <- function(x){
  if (x < 0.001){
    return("< 0.001")
  }
  if (x<0.01 & x>=0.001){
    paste("=", format(x, digits = 3, big.mark = ","))
  }
  else{
    paste("=", format(x, digits = 2, big.mark = ","))
  }
} 

```


#Prepare data
Baseline survival function taken from OPTN policy tables 10-11 & 10-12, pgs. 203-206.  These were taken, imported into an excel and convert to a .csv file for importation into R.  

##Import data
```{r Load_data, message=FALSE}


#Load LAS data
lu_alloc <- read_csv("./Data/lu_alloc.csv")                                  

#Load candidate data
cand_thor <- read_sas("./Data/cand_thor.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

#Load recipient data
tx_lu <- read_sas("./Data/tx_lu.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

#Load diagnosis codes
lu_dx_codes <-read_csv("./Data/Lung Diagnosis.csv") %>% 
  select(group = GROUP, code = CD) %>%
  mutate(group = ifelse((code == 1605), "A", group))                          #Set 1605 as code A for now, will adjust further down

#Load post-transplant baseline survival function
post_surv <- read_csv("./Data/post_transplant_baseline_survival.csv") %>%
  rename(time = Time)                                                         #for variable name consistency 


```

##Select the LAS variables for each patient at the time of transplant
```{r LAS_variables, echo = TRUE, message = FALSE, error=FALSE}

#Find the LAS variables at the time of transplant 
LAS_vars_PT <- lu_alloc %>%
  group_by(PX_ID, CANHX_CHG_DT) %>%
  filter(row_number() == n()) %>%                                             #select the last record for a given day
  group_by(PX_ID) %>%  
  arrange(PX_ID, CANHX_CHG_DT) %>%
  mutate(
    CANHX_CALC_VENTILATOR_USE = ifelse(CANHX_CALC_VENTILATOR_USE == 6 | 
                                         CANHX_CALC_VENTILATOR_USE == 7,
                                       2, CANHX_CALC_VENTILATOR_USE),         #MV coded as 2, 6, and 7 depending on year -> merge
    #fill in creatinine data
    creatinine = ifelse(is.na(CANHX_CALC_SERUM_CREAT), 
                        CANHX_SERUM_CREAT, CANHX_CALC_SERUM_CREAT),           #Fill in with existing LAS data
    max_cr = max(creatinine),                                                 #Estimate rise in creatine
    min_cr = min(creatinine),
    cr_rise = case_when(
      ((max_cr-min_cr)/(min_cr) > 1.5) & max_cr >1  ~ 1,
      TRUE ~ 0 ) )  %>%
   #select the last record for each candidate - this is the one closest to their transplant, if they had a transplant
  slice(n()) %>%   
  ungroup()

```


##Set timeframe and relevant sample
```{r Date_range}

#time frame
start_date <- as.Date("02/19/2015", "%m/%d/%Y")

end_date <- as.Date("02/19/2019", "%m/%d/%Y")
#sample
organ <- "LU"
peds <- TRUE

```


##Clean data - create working sample
```{r Clean_data, echo = TRUE, message = FALSE, error=FALSE}

#Merge data
working_samp_PT <- tx_lu %>% 
  select(-starts_with("DON_")) %>%                                       #drop Donor variables to make the dataset more manageable
  left_join(cand_thor %>% 
              select(PX_ID, CAN_ECMO, CAN_REM_DT, 
                     CAN_DEATH_DT, CAN_AGE_AT_LISTING)) %>%              #Merge with select candidate data
  left_join(LAS_vars_PT %>%select(PX_ID, everything())) %>%              #Merge with LAS variables
  mutate(code = as.numeric(CAN_DGN_LU),
         year = year(REC_TX_DT),
         ECMO = REC_ECMO,
         creatinine = ifelse(is.na(creatinine), 
                             REC_CREAT, creatinine ) ) %>%               #If still issing creatinine, use recipient data
  left_join(lu_dx_codes, by = "code") %>%                                #merge with lung codes
  mutate(
    group = as.character(group),
    group = ifelse((code == 1605 & CANHX_CALC_PAP_MEAN >30), 
                   "D", group))                                          #reassign sarcoid patients to group D if PA pressure > 30


#Filter data and Collect sample sizes for flow chart
print(c("all data", nrow(working_samp_PT)))

working_samp_PT <- working_samp_PT %>%
  filter(REC_TX_DT > start_date & REC_TX_DT <= end_date)              #filter to dates

print(c("exclude out of year", nrow(working_samp_PT)))

working_samp_PT <- working_samp_PT %>%  
  filter(ORG_TY == organ)                                                #filter to lung transplant   
print(c("exclude non-lung TX", nrow(working_samp_PT)))

working_samp_PT <- working_samp_PT %>% 
  filter(CAN_AGE_AT_LISTING >=12)	                                       #filter to recipients listed > 12 years
print(c("exclude >=12", nrow(working_samp_PT)))  



#Create one death variable
working_samp_PT <- working_samp_PT %>% 
   mutate(death_dt = as.Date(pmax(PERS_SSA_DEATH_DT, PERS_OPTN_DEATH_DT, 
                                  PERS_RESTRICT_DEATH_DT, TFL_DEATH_DT, 
                                  na.rm = TRUE )))  


#remove pediatric candidates if indicated
if (peds == FALSE){
	working_samp_PT <- working_samp_PT %>% filter(REC_AGE_AT_TX >17)	
}


initial_sample_size_PT <- nrow(working_samp_PT)


```
The initial sample size is `r initial_sample_size_PT`


#Explore Data - to check for data quality

```{r}
working_samp_PT %>%
  group_by(year(REC_TX_DT)) %>%
  count()
```


```{r Explore_data_means}

#Find the mean value of each variable used in the post_xb calculation, by transplant year
working_samp_PT %>%
  group_by(year(REC_TX_DT)) %>%
  summarise(REC_AGE_AT_TX = mean(REC_AGE_AT_TX),
            creatinine = mean(creatinine, na.rm=TRUE),
            cr_rise = mean(cr_rise, na.rm = TRUE),
            CANHX_CI = mean(CANHX_CI, na.rm = TRUE),
            CANHX_CALC_VENTILATOR_USE = mean(CANHX_CALC_VENTILATOR_USE, na.rm = TRUE),
            CANHX_CALC_O2_AMT = mean(CANHX_CALC_O2_AMT, na.rm=TRUE),
            CANHX_FUNCTN_STAT = mean(CANHX_FUNCTN_STAT, na.rm=TRUE),
            CANHX_CALC_SIX_MIN_WALK = mean(CANHX_CALC_SIX_MIN_WALK, na.rm=TRUE)
        )

```

```{r Explore_data_miss}

#Find the percent of individuals missing values for each post_xb variable, by transplant year
working_samp_PT %>%
  group_by(year(REC_TX_DT)) %>%
  summarise(miss_AGE = mean(is.na(REC_AGE_AT_TX)),
            creatinine = mean(is.na(creatinine)),
            miss_cr_rise = mean(is.na(cr_rise)),
            miss_CI = mean(is.na(CANHX_CI)),
            miss_VENT = mean(is.na(CANHX_CALC_VENTILATOR_USE)),
            miss_O2 = mean(is.na(CANHX_CALC_O2_AMT)),
            miss_FUNCTN = mean(is.na(CANHX_FUNCTN_STAT)),
            miss_SIX = mean(is.na(CANHX_CALC_SIX_MIN_WALK))
  )

```


#Post-transplant Xbeta calculation

"If values for certain covariates are missing, expired, or below the threshold as defined by Table 10- 1, then the LAS calculation will substitute normal or least beneficial values to calculate the candidate's LAS. A normal value is one that a healthy individual is likely to exhibit. A least beneficial value is one that will calculate the lowest LAS for a candidate. Table 10-1 lists the normal and least beneficial values that will be substituted."

below is the code that sums the appropriate waitlist xbeta coefficents.  The coefficents are the 16 digit coefficents used in the OPTN_policy.pdf https://optn.transplant.hrsa.gov/media/1200/optn_policies.pdf.

```{r Sum_of_post_x_betas}

working_samp_PT <- working_samp_PT %>% mutate(
  post_xb = 
    case_when(
      REC_AGE_AT_TX < 46 ~ 0,         
      TRUE ~ 0.0246579831271869*(REC_AGE_AT_TX - 45.9972602)) +      
    case_when(
      is.na(creatinine) & REC_AGE_AT_TX >= 18 ~ 0.0895569900508900*40,   
      REC_AGE_AT_TX >= 18 ~ 0.0895569900508900*creatinine,
      TRUE ~ 0) +
    case_when(
      cr_rise == 1 ~ 0.7708616024698100,
      TRUE ~ 0) +
    case_when(
      CANHX_CI < 2 ~ 0.3499381679822400,
      TRUE ~ 0) +
    case_when(
      ECMO == 1 ~ 0.6094478988424900,
      CANHX_CALC_VENTILATOR_USE == 2 | is.na(CANHX_CALC_VENTILATOR_USE) ~ 0.6094478988424900,   
      TRUE ~ 0) + 
    case_when(
      group == "A" ~ 0,
      group == "B" ~ 0.6115547319209300,
      group == "C" ~ 0.3627014422464200,
      group == "D" ~ 0.4641392063023200) +
    case_when(
      #bronchectasis
      group == "A" & code == 1608 ~ 0.1889100379099400,
      #eisenmonger
      group == "B" & code %in% c(1500, 1548, 1549, 1502, 1501) ~ 0.9146727886744700,
      #LAM
      group == "A" & code == 1611 ~ -1.5194416206749400,
      #OB (no re-tx)
      group == "D" & code == 1612 ~ -1.2050508750702600,
      #pulmonary fibrosis, not IPF
      group == "D" & code == 1613 ~ -0.0723596761367600,
      #sarcoid w/ ph qualifiers
      group == "D" & code == 1605 & CANHX_CALC_PAP_MEAN > 30 ~ -0.0437880049066331,
      group == "A" & code == 1605 & CANHX_CALC_PAP_MEAN <= 30 ~ -0.1389363636019300,
      group == "A" & code == 1605 & is.na(CANHX_CALC_PAP_MEAN) ~ -0.1389363636019300,   
      TRUE ~ 0)+
    case_when(
      ECMO == 1 & group == "A" ~ 26.33*0.0747978926517300,
      ECMO == 1 ~ 26.33* 0.0164276945879309,
      is.na(CANHX_CALC_O2_AMT) & group == "A" ~ 26.33*0.0747978926517300,
      group == "A" ~ CANHX_CALC_O2_AMT*0.0747978926517300,     
      is.na(CANHX_CALC_O2_AMT) ~ 26.33*0.0164276945879309,
      TRUE ~ CANHX_CALC_O2_AMT*0.0164276945879309) +
    case_when(
      CANHX_FUNCTN_STAT == 1 ~ -0.1900086366785100,
      TRUE ~ 0) +
    case_when(
      CANHX_CALC_SIX_MIN_WALK > 1200 ~ 0,
      !is.na(CANHX_CALC_SIX_MIN_WALK) ~ 0.0004594953809594*(1200-CANHX_CALC_SIX_MIN_WALK),  
      TRUE ~ 0.0004594953809594*(1200-0))
  )

#Find how many missing post xb
missing_xb_PT <- working_samp_PT %>%
  filter(is.na(post_xb)) %>%
  nrow()

#Remove those missing post xb
working_samp_PT<- working_samp_PT %>%
  filter(!is.na(post_xb))         

print(c("after remove missing post xbeta", nrow(working_samp_PT)))


```


```{r Histogram_xb}

hist(working_samp_PT$post_xb) 

```



# Waitlist Xbeta calculation

"If values for certain covariates are missing, expired, or below the threshold as defined by Table 10- 1, then the LAS calculation will substitute normal or least beneficial values to calculate the candidate's LAS. A normal value is one that a healthy individual is likely to exhibit. A least beneficial value is one that will calculate the lowest LAS for a candidate. Table 10-1 lists the normal and least beneficial values that will be substituted."

below is the code that sums the appropriate waitlist xbeta coefficents.  The coefficents are the 16 digit coefficents used in the OPTN_policy.pdf https://optn.transplant.hrsa.gov/media/1200/optn_policies.pdf. 

```{r Sum_of_wait_x_betas}

working_samp_PT <- working_samp_PT %>% mutate(
   wait_xb = 
   0.0083990318885565*REC_AGE_AT_TX +                                        #Though this is WL_xb, use current age (ie age at tx)
    case_when(
      CANHX_CALC_BMI < 20 ~ 0.1261444133358100*(20 - CANHX_CALC_BMI),
      TRUE ~ 0)+
    case_when(
      CANHX_CI < 2 ~ 0.5435368888028200,
      TRUE ~ 0)+
    case_when(
      group == "B" & CANHX_CVP_INT > 7 ~ 0.0173841981251578*(CANHX_CVP_INT - 7),
      TRUE ~ 0) +
    case_when( 
      ECMO == 1 ~ 1.6771121096052300, 
      CANHX_CALC_VENTILATOR_USE == 2 ~ 1.6771121096052300,
      TRUE ~ 0) +
    case_when(
       REC_AGE_AT_TX  >= 18 & !is.na(creatinine) ~ 0.5034346761960600*creatinine,
       REC_AGE_AT_TX >= 18 ~ 0.5034346761960600*(0.1),
       TRUE ~ 0) +
    case_when(
      CANHX_CALC_DIAB %in% c("I", "N") ~ 0.4680254026735700,
      TRUE ~ 0) + 
   case_when(
      group == "A" ~ 0,
      group == "B" ~ 1.5774243292137200,
      group == "C" ~ 1.2313926484343600,
      group == "D" ~ 0.6259577164157700) +
    case_when(
      #bronchectasis
      group == "A" & code == 1608 ~ 0.6680518055684700,
      #eisenmonger
      group == "B" & code %in% c(1500, 1548, 1549, 1502, 1501) ~ -0.6278657824830000,
      #LAM
      group == "A" & code == 1611 ~ -0.3162937838984600,
      #OB (no re-tx)
      group == "D" & code == 1612 ~ 0.4453284411081100,
      #pulmonary fibrosis, not IPF 
      group == "D" & code == 1613 ~ -0.2091170018125500,
      #sarcoid
      group == "D" & code == 1605 & CANHX_PAP_MEAN > 30 ~ -0.4577749354638600,
      group == "A" & code == 1605 & CANHX_PAP_MEAN <= 30 ~ 0.9330846239906700,
      TRUE ~ 0) +
    case_when(
     group == "D" & CANHX_CALC_FVC_PREDICTED < 80 ~ 0.1829476350587400*(80-CANHX_CALC_FVC_PREDICTED)/10,
      TRUE ~ 0) + 
    case_when(
      CANHX_FUNCTN_STAT == 1 ~ -0.4471034284458400,
      is.na(CANHX_FUNCTN_STAT) ~  -0.4471034284458400,
      TRUE ~ 0)+
     case_when(
       ECMO == 1 & group == "B" ~  0.0213187586203456*26.33,
       ECMO == 1 ~  0.1188479817592500*26.33,
       is.na(CANHX_CALC_O2_AMT) ~ 0,
       group == "B" ~ 0.0213187586203456*CANHX_CALC_O2_AMT,
       TRUE ~ 0.1188479817592500*CANHX_CALC_O2_AMT) +
    case_when(
      is.na(CANHX_CALC_PCO2) | CANHX_CALC_PCO2 < 40 ~ 0.1104609835819100*(40/10),
      CANHX_CALC_PCO2 >= 40 ~ 0.1104609835819100*(CANHX_CALC_PCO2/10),
      TRUE ~ 0) +
    case_when(
      CANHX_PCO2_THRESHOLD_CHG == 1 ~ 0.2331149280428300,
      TRUE ~ 0) +
    case_when(
      CANHX_CALC_PAP_SYST > 40 & group == "A" ~ 0.4155116686114300*(CANHX_CALC_PAP_SYST -40)/10,
      group == "A" ~ 0,
      CANHX_CALC_PAP_SYST < 20 | is.na(CANHX_CALC_PAP_SYST) ~ 0.0462410402627318*(20)/10,
      TRUE ~ 0.0462410402627318*(CANHX_CALC_PAP_SYST)/10) +
    case_when( 
      !is.na(CANHX_CALC_SIX_MIN_WALK)  ~ -0.0844896372724000*CANHX_CALC_SIX_MIN_WALK/100,
      TRUE ~ -0.0844896372724000*(4000)/100
      )
)

#Find how many missing wait xb
missing_wxb_PT <- working_samp_PT %>%
  filter(is.na(wait_xb)) %>%
  nrow()

#Remove those missing wait xb
working_samp_PT <- working_samp_PT %>%
  filter(!is.na(wait_xb))         

print(c("after remove missing wait xbeta", nrow(working_samp_PT)))


```


#Create death and censor variables for observed survival analysis

Per the original LAS paper (Egan 2006 AJT)

>**Post transplant mortality**
To examine the impact of candidate characteristics on posttransplant mortality within 1 year, a multivariable logistic regression analysis was performed. For patients with known status at 1 year (alive or dead), a weight of 1 was used. For patients reported as alive but with incomplete followup at 1 year, a weight was used that corresponded to the proportion of the 365-day interval for which the patient's status was known. All factors collected on the OPTN TCR and Transplant Recipient Registration forms that were not considered manipulable and were deemed appropriate for the use in an allocation system were considered as potential predictors for posttransplant mortality. This included a wide variety of demographic, clinical, hemodynamic, and pulmonary function factors. The cohort for this analysis included all deceased donor lung-only transplants performed between January 1, 1996 and June 30, 1999, with COPD (n = 1422), CF (n = 498), IPF (n = 463) or PPH (n = 146). There were no exclusions for recipient age or previous transplant.

```{r Create_time_variables}

working_samp_PT<- working_samp_PT %>% 
  filter(!is.na(TFL_LAFUDATE) & !is.na(REC_TX_DT)) %>%        #removing recipients without valid folow-up or tx dates
  mutate(
  censor_range = TFL_LAFUDATE - REC_TX_DT,                    #Number of days patient in sample after their tx
  death_range = death_dt - REC_TX_DT,                         #Number of days patient in sample from tx to death
  death1yr = case_when(                                       #Death within 1 year if die within one year of tx
    TFL_LASTATUS == "D" & death_range <= 364  ~ 1, 
    TRUE ~ 0),     
  time = as.numeric(case_when(                                #Number of days patient is in the sample, whether death or censor, bound at 364
    death1yr == 1 ~ death_range,
    censor_range <= 364 ~ censor_range,
    TRUE ~ 364))                                              #note: day 1 is coded as 0
  ) %>%
  filter(time >= 0)
 
 
print(c("after remove time inconsistencies", nrow(working_samp_PT)))

```




```{r Plot_distribution_failure_times, message = FALSE, warning = FALSE}

#Failure time distribution (censored at one year)
ggplot(working_samp_PT) +
  geom_histogram(aes(x = time))

```




#Export data
```{r Export}

write_csv(working_samp_PT, path = "./Data/Posttransplant.csv")


```


