---
title: "Waitlist Data Clean"
output:
    html_notebook:
      toc: yes
      toc_depth: 2
      toc_float: yes

---
 
  PROJECT:    	Accuracy of of LAS
  
  PROGRAMMERS: 	Nicole Dussault, William Parker
  
  PROGNAME:   	Waitlist_Data_Clean
  
  CREATED:    	7/31/2019
  
  LAST UPDATED: "`r format(Sys.time(), '%d %B, %Y')`"
  
  OBJECTIVE: 	  Create Waitlist dataset for LAS accuracy analysis 
  
  NOTES:        Updated for Feb 2015 - Feb 2019
  

```{r Load_libraries, include = FALSE}

rm(list=ls())
library(tidyverse)
library(haven)
library(lubridate)
library(survival)
library(ggfortify)



```

```{r Convenience_functions, include = FALSE}

#Convenience functions

numextract <- function(string){
  as.numeric(str_extract(string, "\\-*\\d+\\.*\\d*"))
}

comma <- function(x) format(x, digits = 3, big.mark = ",")

comma_0 <- function(x) format(x, digits = 0, big.mark = ",")


comma_1 <- function(x) format(x, digits = 1, big.mark = ",")

comma_2 <- function(x) format(x, digits = 2, big.mark = ",")

comma_p <- function(x){
  if (x < 0.001){
    return("< 0.001")
  }
  if (x<0.01 & x>=0.001){
    paste("=", format(x, digits = 3, big.mark = ","))
  }
  else{
    paste("=", format(x, digits = 2, big.mark = ","))
  }
} 

```

# Prepare data
Baseline survival function taken from OPTN policy tables 10-11 & 10-12, pgs. 203-206.  These were taken, imported into an excel and convert to a .csv file for importation into R.  

## Import data
```{r Load_data, message=FALSE}
                                   
#Load LAS data
lu_alloc <- read_csv("./Data/lu_alloc.csv")

#Load candidate data
cand_thor <- read_sas("./Data/cand_thor.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

#Load recipient data
tx_lu <- read_sas("./Data/tx_lu.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

#Load diagnosis codes
lu_dx_codes <-read_csv("./Data/Lung Diagnosis.csv") %>% 
  select(group = GROUP, code = CD) %>%
  mutate(group = ifelse((code == 1605), "A", group))                         #Set 1605 as code A for now, will adjust further down

#Load waitlist baseline survival function
wait_surv <- read_csv("./Data/wait_list_baseline_survival.csv") %>%
  rename(time = Time)                                                         #for variable name consistency 


```

## Select the LAS variables for each patient at the time of listing
```{r LAS_variables, echo = TRUE, message = FALSE, error=FALSE}

#Find the LAS variables at the time of listing.
LAS_vars_WL <- lu_alloc %>%
  group_by(PX_ID) %>%
  mutate(first_day = min(CANHX_CHG_DT)) %>%                                 #filter to only the first day forms
  filter(CANHX_CHG_DT == first_day) %>%
  ungroup() %>%
  group_by(PX_ID) %>%
  filter(row_number()== n()) %>%                                            #select the last record on the first day
  mutate(                                     
    CANHX_CALC_VENTILATOR_USE = ifelse(CANHX_CALC_VENTILATOR_USE == 6 |     #MV coded as 2, 6, and 7 depending on year -> merge
                                         CANHX_CALC_VENTILATOR_USE == 7, 
                                       2, CANHX_CALC_VENTILATOR_USE),
    #fill in creatinine data
    creatinine = ifelse(is.na(CANHX_CALC_SERUM_CREAT), 
                        CANHX_SERUM_CREAT, CANHX_CALC_SERUM_CREAT)         #Fill in with existing LAS data
    ) %>%
  ungroup()



```


## Set timeframe and relevant sample
```{r Date_range}

#time frame
start_date <- as.Date("02/19/2015", "%m/%d/%Y")

end_date <- as.Date("02/19/2019", "%m/%d/%Y")
#sample
organ <- "LU"
peds <- TRUE

```



## Clean data - create working sample
```{r Clean_data, echo = TRUE, message = FALSE, error=FALSE}

#Merge data
working_samp_WL <- cand_thor %>%
  left_join(LAS_vars_WL %>%select(PX_ID, everything())) %>%               #Merge with LAS variables
  mutate(code = as.numeric(CAN_DGN_LU),
         year = year(CAN_LISTING_DT),
         ECMO = CAN_ECMO) %>% 
  left_join(lu_dx_codes, by = "code") %>%                                 #merge with lung codes
  mutate(
    group = as.character(group),
    group = ifelse((code == 1605 & CANHX_CALC_PAP_MEAN >30), 
                   "D", group),                                           #reassign sarcoid patients to group D if PA pressure > 30
   CANHX_CALC_BMI = ifelse(CANHX_CALC_BMI > 100, NA, CANHX_CALC_BMI)      #Reassign BMI to missing if > 100
  )

#Filter data and Collect sample sizes for flow chart
print(c("all data", nrow(working_samp_WL)))

working_samp_WL <- working_samp_WL %>%
  filter(CAN_LISTING_DT > start_date & CAN_LISTING_DT <= end_date)        #filter to dates

print(c("exclude out of year", nrow(working_samp_WL)))

working_samp_WL <- working_samp_WL %>%  
  filter(WL_ORG == organ)                                                 #filter to lung transplant   
print(c("exclude non-lung TX", nrow(working_samp_WL)))

working_samp_WL <- working_samp_WL %>% 
  filter(CAN_AGE_AT_LISTING >=12)	                                        #filter to recipients listed > 12 years
print(c("exclude >=12", nrow(working_samp_WL)))  


#Create one death variable 
working_samp_WL <- working_samp_WL %>% 
   mutate(death_dt = as.Date(pmax(PERS_SSA_DEATH_DT, PERS_OPTN_DEATH_DT, 
                                  PERS_RESTRICT_DEATH_DT, na.rm = TRUE))) 


#remove pediatric candidates if indicated
if (peds == FALSE){
	working_samp_WL <- working_samp_WL %>% filter(CAN_AGE_AT_LISTING >17)	
}


initial_sample_size_WL <- nrow(working_samp_WL)

```
The initial sample size is `r initial_sample_size_WL`


# Explore Data - to check for data quality

```{r}
working_samp_WL %>%
  group_by(year(CAN_LISTING_DT)) %>% 
  count()
```



```{r Explore_data_means}

#Find the mean value of each variable used in the wait_xb calculation, by listing year
working_samp_WL %>%
  group_by(year(CAN_LISTING_DT)) %>%
  summarise(CAN_AGE_AT_LISTING = mean(CAN_AGE_AT_LISTING),
            CANHX_CALC_BMI = mean(CANHX_CALC_BMI, na.rm=TRUE),
            CANHX_CI = mean(CANHX_CI, na.rm = TRUE),
            CANHX_CVP_INT = mean(CANHX_CVP_INT, na.rm = TRUE),
            CANHX_CALC_VENTILATOR_USE = mean(CANHX_CALC_VENTILATOR_USE, na.rm = TRUE),
            creatinine = mean(creatinine, na.rm=TRUE),
            CANHX_PAP_MEAN = mean(CANHX_PAP_MEAN, na.rm=TRUE),
            CANHX_CALC_FVC_PREDICTED = mean(CANHX_CALC_FVC_PREDICTED, na.rm=TRUE),
            CANHX_FUNCTN_STAT = mean(CANHX_FUNCTN_STAT, na.rm=TRUE),
            CANHX_CALC_O2_AMT = mean(CANHX_CALC_O2_AMT, na.rm=TRUE),
            CANHX_CALC_PCO2 = mean(CANHX_CALC_PCO2, na.rm=TRUE),
            CANHX_PCO2_THRESHOLD_CHG = mean(CANHX_PCO2_THRESHOLD_CHG, na.rm=TRUE),
            CANHX_CALC_PAP_SYST = mean(CANHX_CALC_PAP_SYST, na.rm=TRUE),
            CANHX_CALC_SIX_MIN_WALK = mean(CANHX_CALC_SIX_MIN_WALK, na.rm=TRUE)
        )
```


```{r Explore_data_miss}

#Find the percent of individuals missing values for each wait_xb variable, by listing year
working_samp_WL %>%
  group_by(year(CAN_LISTING_DT)) %>%
  summarise(miss_AGE = mean(is.na(CAN_AGE_AT_LISTING)),
            miss_BMI = mean(is.na(CANHX_CALC_BMI)),
            miss_CI = mean(is.na(CANHX_CI)),
            miss_CVP = mean(is.na(CANHX_CVP_INT)),
            miss_VENT = mean(is.na(CANHX_CALC_VENTILATOR_USE)),
            miss_CR = mean(is.na(creatinine)),
            miss_PAP_MEAN = mean(is.na(CANHX_PAP_MEAN)),
            miss_FVC = mean(is.na(CANHX_CALC_FVC_PREDICTED)),
            miss_FUNC = mean(is.na(CANHX_FUNCTN_STAT)),
            miss_O2 = mean(is.na(CANHX_CALC_O2_AMT)),
            miss_PCO2 = mean(is.na(CANHX_CALC_PCO2)),
            miss_PCO2_Thresh = mean(is.na(CANHX_PCO2_THRESHOLD_CHG)),
            miss_PAP_SYST = mean(is.na(CANHX_CALC_PAP_SYST)),
            miss_SIX = mean(is.na(CANHX_CALC_SIX_MIN_WALK))
  )

```

# Waitlist Xbeta calculation

"If values for certain covariates are missing, expired, or below the threshold as defined by Table 10- 1, then the LAS calculation will substitute normal or least beneficial values to calculate the candidate's LAS. A normal value is one that a healthy individual is likely to exhibit. A least beneficial value is one that will calculate the lowest LAS for a candidate. Table 10-1 lists the normal and least beneficial values that will be substituted."

Below is the code that sums the appropriate waitlist xbeta coefficents.  The coefficents are the 16 digit coefficents used in the OPTN_policy.pdf https://optn.transplant.hrsa.gov/media/1200/optn_policies.pdf. 

```{r Sum_of_wait_x_betas}

working_samp_WL <- working_samp_WL %>% mutate(
   wait_xb = 
   0.0083990318885565*CAN_AGE_AT_LISTING +
    case_when(
      CANHX_CALC_BMI < 20 ~ 0.1261444133358100*(20 - CANHX_CALC_BMI),
      TRUE ~ 0)+
    case_when(
      CANHX_CI < 2 ~ 0.5435368888028200,
      TRUE ~ 0)+
    case_when(
      group == "B" & CANHX_CVP_INT > 7 ~ 0.0173841981251578*(CANHX_CVP_INT - 7),
      TRUE ~ 0) +
    case_when( 
      ECMO == 1 ~ 1.6771121096052300, 
      CANHX_CALC_VENTILATOR_USE == 2 ~ 1.6771121096052300,
      TRUE ~ 0) +
    case_when(
       CAN_AGE_AT_LISTING  >= 18 & !is.na(creatinine) ~ 0.5034346761960600*creatinine,
       CAN_AGE_AT_LISTING >= 18 ~ 0.5034346761960600*(0.1),
       TRUE ~ 0) +
    case_when(
      CANHX_CALC_DIAB %in% c("I", "N") ~ 0.4680254026735700,
      TRUE ~ 0) + 
   case_when(
      group == "A" ~ 0,
      group == "B" ~ 1.5774243292137200,
      group == "C" ~ 1.2313926484343600,
      group == "D" ~ 0.6259577164157700) +
    case_when(
      #bronchectasis
      group == "A" & code == 1608 ~ 0.6680518055684700,
      #eisenmonger
      group == "B" & code %in% c(1500, 1548, 1549, 1502, 1501) ~ -0.6278657824830000,
      #LAM
      group == "A" & code == 1611 ~ -0.3162937838984600,
      #OB (no re-tx)
      group == "D" & code == 1612 ~ 0.4453284411081100,
      #pulmonary fibrosis, not IPF 
      group == "D" & code == 1613 ~ -0.2091170018125500,
      #sarcoid
      group == "D" & code == 1605 & CANHX_PAP_MEAN > 30 ~ -0.4577749354638600,
      group == "A" & code == 1605 & CANHX_PAP_MEAN <= 30 ~ 0.9330846239906700,
      TRUE ~ 0) +
    case_when(
     group == "D" & CANHX_CALC_FVC_PREDICTED < 80 ~ 0.1829476350587400*(80-CANHX_CALC_FVC_PREDICTED)/10,
      TRUE ~ 0) + 
    case_when(
      CANHX_FUNCTN_STAT == 1 ~ -0.4471034284458400,
      is.na(CANHX_FUNCTN_STAT) ~  -0.4471034284458400,
      TRUE ~ 0)+
     case_when(
       ECMO == 1 & group == "B" ~  0.0213187586203456*26.33,
       ECMO == 1 ~  0.1188479817592500*26.33,
       is.na(CANHX_CALC_O2_AMT) ~ 0,
       group == "B" ~ 0.0213187586203456*CANHX_CALC_O2_AMT,
       TRUE ~ 0.1188479817592500*CANHX_CALC_O2_AMT) +
    case_when(
      is.na(CANHX_CALC_PCO2) | CANHX_CALC_PCO2 < 40 ~ 0.1104609835819100*(40/10),
      CANHX_CALC_PCO2 >= 40 ~ 0.1104609835819100*(CANHX_CALC_PCO2/10),
      TRUE ~ 0) +
    case_when(
      CANHX_PCO2_THRESHOLD_CHG == 1 ~ 0.2331149280428300,
      TRUE ~ 0) +
    case_when(
      CANHX_CALC_PAP_SYST > 40 & group == "A" ~ 0.4155116686114300*(CANHX_CALC_PAP_SYST -40)/10,
      group == "A" ~ 0,
      CANHX_CALC_PAP_SYST < 20 | is.na(CANHX_CALC_PAP_SYST) ~ 0.0462410402627318*(20)/10,
      TRUE ~ 0.0462410402627318*(CANHX_CALC_PAP_SYST)/10) +
    case_when( 
      !is.na(CANHX_CALC_SIX_MIN_WALK)  ~ -0.0844896372724000*CANHX_CALC_SIX_MIN_WALK/100,
      TRUE ~ -0.0844896372724000*(4000)/100
      )
)

#Find how many missing wait xb
missing_xb_WL <- working_samp_WL %>%
  filter(is.na(wait_xb)) %>%
  nrow()

#Remove those missing wait xb
working_samp_WL <- working_samp_WL %>%
  filter(!is.na(wait_xb))         

print(c("after remove missing wait xbeta", nrow(working_samp_WL)))

```

```{r Histogram_xb}

hist(working_samp_WL$wait_xb) 

```


#Post-transplant Xbeta calculation

Need to also calculate PT xb for waitlist patients in order to calculate their PT score, and thus overall LAS

```{r Sum_of_post_x_betas}

working_samp_WL <- working_samp_WL %>% mutate(
  post_xb = 
    case_when(
      CAN_AGE_AT_LISTING < 46 ~ 0,   #To calculate a post-xb for all candidates at time of listing, need to use age at listing
      TRUE ~ 0.0246579831271869*(CAN_AGE_AT_LISTING - 45.9972602)) +      
    case_when(
      is.na(creatinine) & CAN_AGE_AT_LISTING >= 18 ~ 0.0895569900508900*40,   
      CAN_AGE_AT_LISTING >= 18 ~ 0.0895569900508900*creatinine,
      TRUE ~ 0) +
    #At listing, candidates won't have creatinine rise data
    #case_when(
     # cr_rise == 1 ~ 0.7708616024698100,
      #TRUE ~ 0) +
    case_when(
      CANHX_CI < 2 ~ 0.3499381679822400,
      TRUE ~ 0) +
    case_when(
      ECMO == 1 ~ 0.6094478988424900,
      CANHX_CALC_VENTILATOR_USE == 2 | is.na(CANHX_CALC_VENTILATOR_USE) ~ 0.6094478988424900,   
      TRUE ~ 0) + 
    case_when(
      group == "A" ~ 0,
      group == "B" ~ 0.6115547319209300,
      group == "C" ~ 0.3627014422464200,
      group == "D" ~ 0.4641392063023200) +
    case_when(
      #bronchectasis
      group == "A" & code == 1608 ~ 0.1889100379099400,
      #eisenmonger
      group == "B" & code %in% c(1500, 1548, 1549, 1502, 1501) ~ 0.9146727886744700,
      #LAM
      group == "A" & code == 1611 ~ -1.5194416206749400,
      #OB (no re-tx)
      group == "D" & code == 1612 ~ -1.2050508750702600,
      #pulmonary fibrosis, not IPF
      group == "D" & code == 1613 ~ -0.0723596761367600,
      #sarcoid w/ ph qualifiers
      group == "D" & code == 1605 & CANHX_CALC_PAP_MEAN > 30 ~ -0.0437880049066331,
      group == "A" & code == 1605 & CANHX_CALC_PAP_MEAN <= 30 ~ -0.1389363636019300,
      group == "A" & code == 1605 & is.na(CANHX_CALC_PAP_MEAN) ~ -0.1389363636019300,   #!!!
      TRUE ~ 0)+
    case_when(
      ECMO == 1 & group == "A" ~ 26.33*0.0747978926517300,
      ECMO == 1 ~ 26.33* 0.0164276945879309,
      is.na(CANHX_CALC_O2_AMT) & group == "A" ~ 26.33*0.0747978926517300,
      group == "A" ~ CANHX_CALC_O2_AMT*0.0747978926517300,     
      is.na(CANHX_CALC_O2_AMT) ~ 26.33*0.0164276945879309,
      TRUE ~ CANHX_CALC_O2_AMT*0.0164276945879309) +
    case_when(
      CANHX_FUNCTN_STAT == 1 ~ -0.1900086366785100,
      TRUE ~ 0) +
    case_when(
      CANHX_CALC_SIX_MIN_WALK > 1200 ~ 0,
      !is.na(CANHX_CALC_SIX_MIN_WALK) ~ 0.0004594953809594*(1200-CANHX_CALC_SIX_MIN_WALK),  
      TRUE ~ 0.0004594953809594*(1200-0))
  )

#Find how many missing post xb
missing_pxb_WL <- working_samp_WL %>%
  filter(is.na(post_xb)) %>%
  nrow()

#Remove those missing post xb
working_samp_WL<- working_samp_WL %>%
  filter(!is.na(post_xb))        

print(c("after remove missing post xbeta", nrow(working_samp_WL)))


#Compare wait xb to post xb
ggplot(data=working_samp_WL, aes(x=wait_xb, y=post_xb)) +
    geom_point()


```



# Create death and censor variables for observed survival analysis

Per the original LAS paper (Egan 2006 AJT)

>**Waiting list mortality**
Mortality on the waiting list was examined using a Cox proportional hazards model. Candidates removed from the waiting list for transplant or reasons other than death were censored at the time of removal from the waiting list. The chronologic time from listing was used, regardless of whether a patient was inactive on the waiting list at any point during this period, as candidates could have died while inactive. Initially, all factors collected on the OPTN Transplant Candidate Registration (TCR) form at the time
of listing were considered as potential predictors of waiting list mortality. This included a wide variety of demographic, medical and social history, clinical, hemodynamic, and pulmonary function factors. 

```{r Create_time_variables}

#Treat delisting and transplant as a censoring event, so only deaths on the waitlist count
working_samp_WL <- working_samp_WL %>% 
  filter(!is.na(CAN_LISTING_DT)) %>%               #removing candidates without valid listing dates.
  mutate(
  censor_range = CAN_REM_DT - CAN_LISTING_DT,      #Number of days patient is in the sample from listing to removal
  death_range = death_dt - CAN_LISTING_DT,         #Number of days patient is in the sample from listing to death
  death1yr = case_when(                            #Death within 1 year if Removal code is death and die within one year 
    CAN_REM_CD == 8 & death_range < 365 ~1, 
    TRUE ~0),
  time = case_when(                                #Number of days patient is in the sample, whether death or censor, bound at 364
    death1yr == 1 ~ death_range,
    censor_range < 365 ~ as.numeric(censor_range),
    TRUE ~ 364)                                    #note: day 1 is coded as 0
  )%>%
  filter(time >= 0) 

print(c("after remove time inconsistencies", nrow(working_samp_WL)))

#Sensitivity - create separate death variables treating death after removal for too sick (rem code 13) as death on waitlis
working_samp_WL <- working_samp_WL %>% 
  mutate(
    death1yr2 = case_when(
      CAN_REM_CD == 8 & death_range < 365 ~ 1,
      CAN_REM_CD == 13 & death_range < 365 ~1,
      TRUE ~0),
    time2 = case_when(
      death1yr2 == 1 ~ as.numeric(death_range),
      censor_range < 365 ~ as.numeric(censor_range),
      TRUE ~ 364),
    time2 = ifelse(time2 > 364, 364, time2) ) %>%
  filter(time2 >= 0 & time2 <= 364) 


print(c("after remove time inconsistencies, second death definition", nrow(working_samp_WL)))


```



```{r Plot_distribution_failure_times, message = FALSE, warning = FALSE}

#Failure time distribution (censored at one year or delisting/transplant)
ggplot(working_samp_WL) +
  geom_histogram(aes(x = time))


```




# Export data
```{r Export}

write_csv(working_samp_WL, path = "./Data/Waitlist.csv")

```


